# Imputation - what is it good for?

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

## Overview 

A common refrain in scRNA-seq method development is the need to impute zeros due to dropout events.
The idea is to reduce noise by removing technical zeroes, thus improving the resolution of downstream analysis.
The fundamental difficulty here is that you don't know whether a zero count is a true dropout or due to a genuine lack of expression.
This is hard to ascertain for any given observation - you can only estimate the proportion of dropouts for a gene, and even then that requires spike-ins.

## Issues with existing methods 

Effective imputation requires knowledge of the biological structure in the dataset.
We can make some reasonable assumptions about the distribution of counts due to technical noise for cells from the same population.
This avoids trying to impute across different subpopulations of cells, some of which may be genuinely silent in their expression of a particular gene.

We can obtain the structure from the data, based the similarities between the cellular expression profiles.
The problem is that the imputation is also done on the same data, resulting in some circularities when the imputed data are used for defining structure.
This is most clearly demonstrated by generating unstructured data, and seeing how the imputation methods subsequently introduce and reinforce any structure that might be present.
For example, using [MAGIC](https://github.com/KrishnaswamyLab/MAGIC):

```{r}
set.seed(200)
counts <- matrix(rpois(100000, 10), ncol=100)

# devtools::install_github("KrishnaswamyLab/MAGIC/R") 
# last tested 1be8cbfbe645331a29e4867ae84a3344d368b45c
library(Rmagic) 

# Guessing that rows are cells, based on its use of rowSums() 
# to compute library size (seriously?)
mgc <- run_magic(t(counts))

# Creating a PCA plot.
library(scater)
sce <- SingleCellExperiment(list(imputed=t(mgc)))
sce <- normalize(sce, exprs_values="imputed") 
plotPCA(sce) # >80% variance explained!

# In contrast, without any imputation:
sce2 <- SingleCellExperiment(list(counts=counts))
sce2 <- normalize(sce2)
plotPCA(sce2) # 2% variance explained.
```

Or using [_scImpute_](https://github.com/Vivianstats/scImpute).
Note that I've used a lower Poisson mean here, as _scImpute_ seems to only do something when there are zeros.

```{r}
set.seed(100)
counts <- matrix(rpois(100000, 1), ncol=100)

# devtools::install_github("Vivianstats/scImpute")
# last tested 98b78b5b044ade682d75930eca4d936d1d85d615
library(scImpute) 

# Need to save counts to file (!!)
xpath <- tempdir()
count_path <- file.path(xpath, "counts.csv")
rownames(counts) <- paste0("Gene", seq_len(nrow(counts)))
colnames(counts) <- paste0("Cell", seq_len(ncol(counts)))
write.csv(counts, file=count_path)

# Need '/' as they don't use file.path (!!!)
sci <- scimpute(count_path, out_dir=paste0(xpath, "/"), Kcluster=2)
imputed <- read.csv(file.path(xpath, "scimpute_count.csv"), 
    header=1, row.names=1)

# Creating a PCA plot.
library(scater)
sce <- SingleCellExperiment(list(imputed=as.matrix(imputed)))
sce <- normalize(sce, exprs_values="imputed") 
plotPCA(sce) # Woah, what a nice structure.

# In contrast, without any imputation:
sce2 <- SingleCellExperiment(list(counts=counts))
sce2 <- normalize(sce2)
plotPCA(sce2) # 2% variance explained.
```

A generous interpretation of these results would be that current imputation schemes operate more like smoothers, reducing noise in the plot to highlight existing structure.
But really, if I wanted prettier pictures, I would just shrink all points to their cluster centers - job done.

## Further thoughts

One _could_ imagine an approach whereby you distinguish between biological and technical noise with spike-ins.
This will allow you to determine, on probability, whether a zero is a dropout or not.
You could then obtain a posterior distribution for each observation, and use the MAP estimate as the imputed value for zeroes.
This brings us very close to the "denoised expression values" computed by _BASiCS_ or _BISCUIT_.
However, _BASiCS_ assumes all cells are in a single group, and fails to account for substructure that can affect the expected count (and thus the probability of dropout).
_BISCUIT_ allows for some substructure with a Dirichlet process, but assumes normality that is definitely not present.
Both would probably produce distorted MAP estimates that are of questionable benefit.

So, there is no imputation without denoising, given that the zeroes are just another consequence of technical noise.
Our approach is to handle the drop-outs on a per-gene basis, by modelling the technical variability introduced by them.
This is simpler, involves fewer assumptions, and with the low-rank reconstructions from PCA, we can also obtain denoised expression values directly.
