---
title: Some thoughts on visualization
author: Aaron Lun
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
bibliography: ref.bib
---   

```{r, echo=FALSE, message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
set.seed(120000)
```

# Comments on dot plots

For anyone who doesn't know what I'm talking about, it's the `r CRANpkg("Seurat")`-style dot plots here:

```{r, fig.wide=TRUE, fig.asp=0.3}
###########################
### Setting up the data ###
###########################

library(scRNAseq)
sce <- ZeiselBrainData()

library(scran)
library(scater)
sce <- logNormCounts(sce)
markers <- pairwiseTTests(logcounts(sce), sce$level1class)
output <- getTopMarkers(markers[[1]], markers[[2]], n=2)

features <- unlist(unlist(output))
object <- sce
group <- sce$level1class
                                                                        
#################################
### Setting up the statistics ###
#################################
                                                                        
num <- numDetectedAcrossCells(object, ids = group, 
    subset_row = features, average = TRUE)
ave <- sumCountsAcrossCells(object, ids = group, subset_row = features,
    exprs_values="logcounts", average = TRUE)
logfc <- ave - rowMeans(ave)

#####################
### Making a plot ###
#####################

rn <- factor(rownames(logfc))
cn <- factor(colnames(logfc))
evals_long <- data.frame(
    Row = rep(rn, ncol(logfc)),
    Col = rep(cn, each = nrow(logfc)), 
    LogFC = as.numeric(logfc),
    Percent = as.numeric(num)
)

ggplot(evals_long) + 
    geom_point(aes(x=Row, y=Col, colour=LogFC, size=Percent)) +
    scale_color_gradient2(low="blue", high="red")
```

The idea is to use the color to capture the log-fold change from the mean (or any other statistic) while using the size to represent the number of cells with detectable expression.
This allows both statistics to be represented compactly for each gene and cluster in a single plot.
The eye is naturally drawn towards the large red circles, allowing the reader to rapidly focus on candidate markers with strong positive log-fold changes and a high proportion of detected cells.
More importantly, it looks pretty.

However, the dot plot has a few deficiencies as an effective visualization tool.
Many of them stem from the use of the point area to represent... well, anything, really.
Humans are notoriously bad at judging areas (see Tufte, _The Visual Display of Quantitative Information_) so we cannot rely on two dimensions to accurately represent a one-dimensional quantity.
Even worse is that the size of the point actively interferes with visualization of the log-fold change.
We cannot easily see colors for small points, reducing the effectiveness of visualization for genes that are sporadically expressed but critical for cell type identity (e.g., _Cd4_).

A related problem is that the use of size implicitly introduces a secondary color scale involving the background color.
For a low-abundance gene in the above example, the plot transitions from red to grey/white as expression decreases.
If this secondary scale overlaps with the primary color scale, it becomes difficult to interpret, e.g., does a white region of the plot represent a lack of any expression or detectable expression with a zero log-fold change?
The issue arises even if the overlap in color scales is not exact - here, any light color that is similar to the background grey would be bad enough, especially if variations in screen/printer/eye quality are taken into account.

One might argue that all of these issues can be swept aside if we are only interested in identifying high-quality marker genes.
In such cases, readers only need to look for the presence of big red dots and avoid interpreting relative areas or conflicts with the background color.
This is a valid perspective but the presence of the additional colors becomes an unnecessary distraction, e.g., an uninteresting gene/cluster combination can manifest as either a small dot or as a large blue dot.
The reader should not be provided with two different representations that have the same interpretation.

To ensure that readers interpret the dot plot in the "correct" way, we suggest forgoing any attempt to represent the uninteresting statistics faithfully.
Rather, we suggest (i) capping the minimum log-fold change at zero and (ii) synchronizing the zero color with the background color.
Any interpretation of the plot collapses to a simple "is it a big red dot or is it empty?" - there is no need to distinguish between zero log-fold changes and lack of expression, because we do not care to do so.
(Some might complain that this discards information but, as we have discussed, there was no way to visualize that information effectively in the first place.)

# Session information

```{r}
sessionInfo()
```
