---
title: Impact of batch effect removal on PCA
author: Aaron Lun
date: 27 April 2017 
output:
  html_document:
      fig_caption: false 
---

```{r, echo=FALSE, results="hide"}
set.seed(100)
dir.create("figure-batch")
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE, fig.path="figure-batch/")
```

# Overview 

If there are two equivalent batches, one might imagine that you could regress out the batch effect and perform PCA/clustering on the residuals.
This is the approach that is commonly used with `removeBatchEffect` and `comBat`.
However, this is only guaranteed to remove the systematic differences in the batch means.
Some interesting effects arise when zero counts are present.

# Illustration of the problem

Consider the following example:

```{r}
set.seed(20000)
a <- matrix(0, 100, 100)
a[sample(length(a), 100)] <- 1
groupings <- rep(LETTERS[1:2], each=50)
fit <- lm.fit(y=t(a), x=model.matrix(~groupings))
resid <- t(fit$residuals)
out <- prcomp(t(resid))
plot(out$x[,1], out$x[,2], col=c(A="blue", B="red")[groupings])
```

We see a small partition between batches due to the breaking of ties when zero values are converted to residuals.
The residuals corresponding to zeroes will be the same within each batch, but will almost always be different between batches.
Across many genes, this adds up to a systematic difference between batches.

# Possible solutions

The hope would be that this is negligible compared to structure within batches.
Indeed, when we actually have a look at more realistic data that's not 99% zero, the structure disappears.
This provides another case for filtering out low-abundance genes with lots of zeroes.

```{r}
set.seed(20000)
a <- matrix(rpois(10000, lambda=1), 100, 100, byrow=TRUE)
sum(a!=0)/length(a)
fit <- lm.fit(y=t(a), x=model.matrix(~groupings))
resid <- t(fit$residuals)
out <- prcomp(t(resid))
plot(out$x[,1], out$x[,2], col=c(A="blue", B="red")[groupings])
```

... and also when there's a strong batch effect that needs to be removed.
There's no systematic shift between batches, though obviously, you shouldn't interpret the variance here.

```{r}
set.seed(20000)
groupings <- rep(LETTERS[1:2], each=50)
a <- matrix(rpois(10000, lambda=c(A=1, B=5)[groupings]), 100, 100, byrow=TRUE)
fit <- lm.fit(y=t(a), x=model.matrix(~groupings))
resid <- t(fit$residuals)
out <- prcomp(t(resid))
plot(out$x[,1], out$x[,2], col=c(A="blue", B="red")[groupings])
```

One solution would be to setting zero-derived residuals to a constant.
However, this distorts the variance explained by each PC, which is required for `denoisePCA` to work.
The actual value of the constant also matters here - I'm not sure how to choose it.

# Wrapping up

```{r}
sessionInfo()
```
