---
title: Cell cycle classification comments
author: Aaron Lun
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warnings=FALSE)
library(BiocStyle)
```

# Excuses for poor performance

Consider the classifier trained on gene count data for one short gene A and one long gene B.
In G1, B is silent and A is strongly expressed, while in all other phases, B's expression is non-zero (lower molarity than A, but higher CPM due to length).
In read count data, this gene pair would be selected as A > B in G1 and B > A in all other phases.
However, in UMI data, this gene pair would not be informative as A > B in terms of molecule counts in all phases.
I use gene length as an example here but similar effects are observed with mappability, sequenceability, etc.

Another possible contributor to poor performance is the difference in the cells used for training and those in the test data.
If certain expression patterns are associated with the cell cycle in the training set (of mouse embryonic stem cells), these may be incorporated into the classifier.
However, if those patterns are not associated with the cell cycle in the test data, their inclusion will add noise without providing any phase information.
This will lead to a deterioration in the accuracy of the classifier.
In general, this is unlikely to be a major issue as the cell cycle should be a conserved process across many lineages and conditions.

Furthermore, there may be misclassification due to a large number of cells being in G0.
In theory, these should be closest to G1 but they may be different enough that you'd get a low G1 score, making them show up as S-like or even G2M.

# Correcting for cell cycle phase 

An alternative approach uses genes that have annotated functions in cell cycling and division.
We extract all genes associated with the relevant GO terms and use them to construct a PCA plot for the brain dataset.
Figure ((braincyclepca)) contains three clusters that may correspond to distinct phases of the cell cycle.
This can be determined explicitly by identifying marker genes for each cluster as previously described, and checking whether each marker has known phase-specific expression with resources such as [Cyclebase](http://www.cyclebase.org) [@santos2015cyclebase].

```{r, echo=FALSE, results='hide', message=FALSE, eval=FALSE}
library(org.Mm.eg.db)
fontsize <- theme(axis.text=element_text(size=12), axis.title=element_text(size=16))
```

```{r braincyclepca, fig.width=10, fig.height=5, fig.cap="PCA plot of the brain dataset, using only genes with annotated functions in cell cycling or division.", eval=FALSE}
ccgenes <- select(org.Mm.eg.db, keys=c("GO:0022403", "GO:0051301"), keytype="GOALL", column="SYMBOL")
sce <- readRDS("brain_data.rds")
chosen.genes <- which(rownames(sce) %in% ccgenes$SYMBOL)
plotPCA(sce, feature_set=chosen.genes) + fontsize 
```

We can also identify hidden factors of variation across the annotated genes using `r Biocpkg("RUVSeq")`.
This assumes that, if all cells were in the same phase of the cell cycle, there should be no DE across cells for genes associated with the cell cycle.
Any systematic differences between cells are incorporated into the `W` matrix containing the factors of unwanted variation.
These factors can then be included as covariates in the design matrix to absorb cell cycle effects in the rest of the dataset.
We set `k=2` here to capture the variation corresponding to the two principal components in Figure ((braincyclepca)).

```{r, eval=FALSE}
library(RUVSeq)
ruv.out <- RUVg(exprs(sce), isLog=TRUE, cIdx=chosen.genes, k=2)
head(ruv.out$W)
```

In general, we prefer using the `cyclone`-based approach for phase identification and blocking.
This is because the expression of some cell cycle genes may be affected by other biological/experimental factors at the single-cell level, 
much like how the cell cycle affects the expression of non-cell cycle genes in the first place.
As a result, the inferred factors of variation may include interesting differences between cells, such that blocking on those factors would result in loss of detection power.
`cyclone` calls the phase for each cell separately and is more robust to systematic (non-cell-cycle-related) differences between cells.

The obvious example would be if cells in the same phase had different amounts of expression for cell cycle genes, corresponding to some other factor (e.g., treatment).
This would get picked up as a hidden factor of variation, leading to loss of power to detect that other factor.
In contrast, `cyclone` wouldn't care as long as the relative amounts of expression of cell cycle genes (or more robustly, their ranked expression) within each cell are unchanged.
Using `scLVM` would also work as it would learn the interesting factors of variation and block on them to avoid them getting regressed out with cell cycle phase.
However, that kind of model is notoriously finicky due to the large numbers of moving parts.

# Session information

```{r}
sessionInfo()
```
